---
title: "Lab 3a - Exercises"
author: "Environmental Data Analytics | John Fay and Luana Lima "
date: "Fall 2025"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## Set up the session
- Load packaged: tidyverse, lubridate, & here
- Import the `EPAair_PM25_NC2019_raw.csv` file
  - Set strings to be read in as factors
  - Set the "Site.ID" column to be read in as a factor
- Set the date column to be a date

```{r Setup}
#Import packages
library(tidyverse); library(lubridate); library(here)
library(janitor)

#Check workspace
here()

#Import data
pm25 <- read.csv(
  #----ENTER THE FILENAME HERE, USING THE here() PACKAGE----
  here("Data/Raw/EPAair_PM25_NC2019_raw.csv"),
  stringsAsFactors = TRUE,
  colClasses = c("Site.ID"='factor')
) %>% 
  clean_names()

#Fix the date
pm25$date <- mdy(pm25$date)

```


## Explore properties of the dataframe
-   How many different sites are included in the data set?
-   What is the mean daily PM2.5 concentration?
-   In what units of are PM 2.5 concentrations reported?

```{r Explore the data}
# Compute the number of different sites
pm25 %>% distinct(site_id) %>% nrow()

# Compute the mean and median PM 2.5 concentration
names(pm25)
mean(pm25$daily_mean_pm2_5_concentration)    # mean = 7.68
median(pm25$daily_mean_pm2_5_concentration)  # median = 7.4

# What are the units
pm25 %>% distinct(units) %>% pull() 

# What is the date range (min and max dates)
pm25 %>% pull(date) %>% min() # min = Jan 1, 2019
pm25 %>% pull(date) %>% max() # max = Dec 31, 2019


```


## Plot distribution of values

-   Number of samples using each different source 
-   Distribution of daily mean concentration values
-   Distribution of daily mean concentration, by source*
  *   There are multiple ways to visualize this...

```{r Visualize the data}
unique(pm25$source)

#Number of samples using each different source 
ggplot(pm25) + 
  geom_bar(aes(x = source))

#Distribution of daily mean concentration values
ggplot(pm25, aes(x = daily_mean_pm2_5_concentration)) + 
  geom_histogram()

#Distribution of daily mean concentration, by source
ggplot(pm25) + 
  geom_freqpoly(aes(x = daily_mean_pm2_5_concentration, 
                    color = source))


```

