---
title: "Assignment 10: Data Scraping"
author: "Sylvia Hipp"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(rvest)
library(glue)
library(here)

# check wd
here()

my_theme <- theme_classic() + 
  theme(
    legend.position = "bottom"
  )

theme_set(my_theme)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 

website <- read_html(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024")

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 

# function to simplify reading in each element
scrape_html_text <- function(element, website){
  return(
    website %>% 
      html_nodes(element) %>% 
      html_text()
  )
}

# 1. System Information
water_system_name <- "div+ table tr:nth-child(1) td:nth-child(2)" %>% 
  scrape_html_text(website)

pwsid <- "td tr:nth-child(1) td:nth-child(5)" %>% 
  scrape_html_text(website)

ownership <- "div+ table tr:nth-child(2) td:nth-child(4)" %>% 
  scrape_html_text(website)

# 3. Water Supply Sources
max_daily_use <- "th~ td+ td" %>% 
  scrape_html_text(website)


```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 

#months <- as.vector(rbind(1:4, 5:8, 9:12))
months <- ".fancy-table:nth-child(30) tr+ tr th" %>% scrape_html_text(website)
year <- 2024

df_water_use <- data.frame("month" = months, 
                           "max_daily_use" = as.numeric(max_daily_use)) %>% 
  mutate(ownership = ownership, 
         pwsid = pwsid, 
         water_system_name = water_system_name, 
         date = ymd(paste(year, month, "1", sep = "-")))

#5 

ggplot(df_water_use, aes(x = date, y = max_daily_use)) + 
  geom_line() + 
  labs(
    x = NULL, 
    y = "Maximum Daily Use (MGD)", 
    title = glue("Maximum Daily Water Use in {water_system_name}, {year}")
  )

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. 

scrape_website <- function(pwsid, year){
  url <- glue(
    "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid={pwsid}&year={year}")
  
  website <- read_html(url)
  
  # scrape relevant data
  water_system_name <- "div+ table tr:nth-child(1) td:nth-child(2)" %>% 
    scrape_html_text(website)
  
  ownership <- "div+ table tr:nth-child(2) td:nth-child(4)" %>% 
    scrape_html_text(website)
  
  max_daily_use <- "th~ td+ td" %>% scrape_html_text(website)
  
  # location of months field changes between years, using another solution
  # months <- ".fancy-table:nth-child(30) tr+ tr th" %>% scrape_html_text(website)
  months <- as.vector(rbind(1:4, 5:8, 9:12))
  
  df_water_use <- data.frame("month" = months, 
                             "max_daily_use" = as.numeric(max_daily_use)) %>% 
    mutate(ownership = ownership, 
           pwsid = pwsid, 
           water_system_name = water_system_name, 
           date = ymd(paste(year, month, "1", sep = "-")))
  
  return(df_water_use)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 

durham_water_use_2020 <- scrape_website(pwsid = "03-32-010", 
                                        year = 2020)

glimpse(durham_water_use_2020)

ggplot(durham_water_use_2020, aes(x = date, y = max_daily_use)) + 
  geom_line() +
  labs(
    x = NULL, 
    y = "Maximum Daily Use (MGD)", 
    title = "Maximum Daily Water Use in Durham, 2020"
  )

```

8. Use the function above to extract data for Cary (PWSID = '03-32-010') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 

cary_water_use_2020 <- scrape_website(pwsid = "03-92-020", year = 2020)

bind_rows(durham_water_use_2020, cary_water_use_2020) %>% 
  ggplot(aes(x = date, y = max_daily_use, color = water_system_name)) + 
  geom_line() + 
  scale_color_brewer(palette = "Set1") +
  labs(
    x = NULL, 
    y = "Maximum Daily Use (MGD)", 
    color = "Water System:",
    title = "Maximum Daily Water Use in Durham and Cary, 2020"
  )

```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023. Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 

cary_water_use_2018_2023 <- c(2018:2023) %>% 
  map(.f = scrape_website, pwsid = "03-92-020") %>% 
  bind_rows()

ggplot(cary_water_use_2018_2023, aes(x = date, y = max_daily_use)) +
  geom_line() + 
  geom_smooth(se = FALSE, method = "loess") +
  labs(
    x = NULL, 
    y = "Maximum Daily Use (MGD)",
    title = "Maximum Daily Water Use in Cary, 2018–2023"
  )
  

```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?

> Answer: 
> The smoothed line suggests there has been a slight positive trend in water usage in Cary between 2018 and 2023. There is also a clear seasonal component to water usage.
