---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Sylvia Hipp"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

**!! AI generated code should NOT be used in this assignment**


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
# Load necessary packages
library(tidyverse)
library(agricolae)
library(here)

# check working directory
here()

# import raw NTL-LTER data file
ntl_lter_chem_physics <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"))

# set sample date to date object
ntl_lter_chem_physics$sampledate <- mdy(ntl_lter_chem_physics$sampledate)

#2
# Build ggplot theme 
mytheme <- theme_classic() + 
  theme(
    # set legend position
    legend.position = "bottom",
    
    # set fonts
    plot.title = element_text(size = 14), 
    axis.title = element_text(size = 12), 
    axis.text = element_text(size = 10, color = "black"),
    legend.text = element_text(size = 10, color = "black")
  ) 

# set custom theme as default
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:

> Answer:
H0: mean lake temperature recorded in July does not change with depth across all lakes
Ha: mean lake temperature recorded in July does change with depth across all lakes


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
ntl_lter_subset <- ntl_lter_chem_physics %>% 
  filter(month(sampledate) == 7) %>% 
  select(lakename, year4, daynum, depth, temperature_C) %>% 
  na.omit()

#5
ggplot(ntl_lter_subset, aes(x = depth, y = temperature_C)) + 
  geom_point(size = 1) + 
  geom_smooth(method = lm) + 
  scale_y_continuous(limits = c(0, 35)) + 
  labs(x = "Depth (m)", 
       y = "Temperature (C)", 
       title = "Relationship Between Lake Depth and Temperature")

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure suggests that there is a negative relationship between temperature and depth. At a certain depth, the temperature levels out around 5 degrees C, which might suggest a non-linear relationship. It could be helpful to plot depth vs. temperature on a logarithmic scale to see if a non-linear relationship fits better.   


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
# regression
temp_depth_regression <- lm(data = ntl_lter_subset, 
                            temperature_C ~ depth)
summary(temp_depth_regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The depth coefficient in the regression is -1.95, which confirms there is a negative correlation: as depth increases 1 meter, we expect temperature to decrease 1.95 degrees C. The intercept is 21.96, which means we expect temperature to be 21.96 degrees C at the surface. The regression has an R^2 of 0.739, which means changes in depth can explain 73.9% of the variability in temperature. These findings are based on 9726 degrees of freedom (9728 observations) and are statistically significant with a p-value < 2.2e-16. 

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
temp_regression_all <- lm(temperature_C ~ year4 + daynum + depth, 
                          data = ntl_lter_subset)
step(temp_regression_all, direction = "backward")


#10

recommended_regression <- lm(temperature_C ~ year4 + daynum + depth, 
                             data = ntl_lter_subset)
summary(recommended_regression)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables suggested by the AIC method is year, date, and depth. This model explains 74.1% of the observed variance in temperature. This is a slight improvement over the model just using depth, which explained 73.9% of variance.


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
# expressed as an ANOVA model
lake_temp_anova <- aov(temperature_C ~ lakename, 
                       data = ntl_lter_subset)
summary(lake_temp_anova)

# expressed as a linear model
lake_temp_anova2 <- lm(temperature_C ~ lakename, 
                       data = ntl_lter_subset)
summary(lake_temp_anova2)


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: the p-value of the ANOVA is < 2.2e-16, which means we can reject our null hypothesis (that there is no difference) and conclude that there is a significant difference in mean temperature among the lakes.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

ggplot(ntl_lter_subset, aes(x = depth, y = temperature_C)) + 
  geom_point(alpha = 0.5, aes(color = lakename)) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  scale_color_brewer(palette = "Paired") +
  scale_y_continuous(limits = c(0, 35)) +
  labs(x = "Depth (m)", 
       y = "Temperature (C)", 
       title = "Relationship Between Lake Depth and Temperature by Lake Name", 
       color = NULL)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(lake_temp_anova)

HSD_groups <- HSD.test(lake_temp_anova, "lakename", group = TRUE)
HSD_groups$groups

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: According to the Tukey's HSD test, assuming an alpha of 0.05, Peter Lake has statistically the same mean temperature as Ward Lake and Paul Lake. 

 
17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: If we are only considering these two lakes, we would be able to use a two-sample t-test. 


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
# filter to just Crampton Lake and Ward Lake
crampton_ward_lake <- ntl_lter_subset %>% 
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

# run t-test
t.test(crampton_ward_lake$temperature_C ~ crampton_ward_lake$lakename)

```

>Answer: The t-test found a p-value of 0.2649, which does not allow us to reject our null hypothesis that the mean July temperature is the same between Crampton Lake and Ward Lake. The sample means are 15.35 degrees C and 14.46 degrees C for Crampton Lake and Ward Lake, respectively. Even though these means are not equal, the t-test indicates that true mean temperatures of these two lakes are statistically the same. This matches the answer for part 16: the Tukey test found a p-value of 0.971 for the pair and assigned both lakes to the same group, so it confirms that these two lakes have statistically the same mean. 
